Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy

### 2025-01-22 manuscript/manuscript.qmd
Suparna and Marc edits

---
title: "Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy"
short-title: "Why Donors Donate"
date: "2025-01-20"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
  orcid: 0000-0003-3506-7590
- name: Marc Dotson
  affiliation: Utah State University
  url: https://occasionaldivergences.com/
  email: marc.dotson@usu.edu
  orcid: 0000-0001-7846-3660
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Governments across the world have increasingly used laws to restrict the work of nonprofits, which has led to a reduction in public or official foreign aid directed towards these groups. Many international nonprofits, in response, have turned to individual donors to offset the loss of traditional fundings. What are individual donors' preferences regarding donating to legally besieged nonprofits abroad? We conducted a conjoint experiment on a nationally representative sample of likely donors in the US and found that learning about host government criticism and restrictions on nonprofits decreases individuals' preferences to donate to them. However, organizational features such as financial transparency and accountability can protect against this dampening effect. Our results have important implications both for understanding private international philanthropy and how nonprofits can better frame their fundraising appeals at a time when they are facing restrictive civic spaces and hostile governments abroad.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/andrewheiss/silent-skywalk>
keywords:
  - philanthropy
  - conjoint experiments
  - donor heuristics
  - repression
  - NGOs
  - civil society
  - nonprofits
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/silent-skywalk>"
bibliography: references.bib
csl: csl/apa.csl
link-citations: true
word-count: "9,300"
---

@@

Non-governmental organizations (NGOs) are a core component of global governance and service delivery. Non-governmental organizations (NGOs) are a core component of global governance and service delivery. Over the last two decades, however, NGOs face restrictive civic spaces abroad [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020]. Repressive governments have increasingly used legal measures to create barriers to funding and advocacy for a variety of NGOs [@DupuyRonPrakash:2016; @Chaudhry:2022; @Heiss:2017]. For instance, in 2012 Russia passed its infamous "Foreign Agent" law (Federal Law No. 121-F2), which forced organizations receiving foreign funding and engaging in political activities to register as foreign agents, which forced hundreds of organizations to close down [@Kirova:2024]. These developments were not limited to authoritarian regimes. Hybrid and democratic states have also restricted NGOs' operations within their territories. Between 2014 and 2019, the Indian government banned 14,500 NGOs from receiving money from abroad, including prominent international groups such as the Ford Foundation and Greenpeace [@PressTrustofIndia:2019]. Policymakers have referred to this phenomenon as the "closing of civic space" or "shrinking civic space" [@CarothersBrechenmacher:2014; @Rutzen:2015b]. Data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in more than 60% of countries---117 in total [@CIVICUSMonitor]. This phenomenon of closing civic space has a large deterrent effect on both advocacy *and* service-delivery NGOs [@SpringmanMaleskyRight:2022].

@@

The amount of private foreign aid is not insignificant---from 2000--2010, private giving by US entities towards international causes doubled in real terms [@DesaiKharas:2018, 505]. In 2018, this amounted to \$22.88 billion in individual donations earmarked specifically for international affairs INGOs [@GivingUSA:2019]. This turn to private donors opens INGOs up to new dynamics in seeking out funds. Individual donor behavior is unlike official donor behavior [@DesaiKharas:2018]. They have different preferences---while official donors channel funds to countries with better institutional quality or to reward governmental performance, [@Bermeo:2011; @DietrichWright:2015], individual donors are more inclined to donate to NGOs in countries experiencing a humanitarian crisis [@DesaiKharas:2018]. As such, individual donors may have separate metrics of effectiveness or impact that can differ from the benchmarks that large donor agencies rely on, and INGOs may need to adapt their fundraising strategies to a different set of potential donors.

While INGOs have control over organizational-level factors that determine their effectiveness, the political environment in the host countries INGOs work in---including the global restrictions on civil society that is causing many organizations to seek out private funding in the first place---is a structural factor that INGOs can rarely influence on their own. How do individual donors in the US feel about donating to legally besieged NGOs abroad? How do organizational elements of private donor decision-making interact with broader structural limits on INGO fundraising? While there are many organizational-level factors that can determine an NGO's impact, including such as organizational structure, financial transparency, accountability, leadership, and strategy [@MitchellSchmitzBruno-vanVijfeijken:2020], in this paper, we focus on the two factors that are most visible to donors outside of the organizations and on which information is available publicly---financial transparency and accountability.

@@

We fielded a conjoint experiment to a nationally representative sample of 1,016 adults in the US. This experiment allows us to assess individual level giving preferences among likely donors in the US towards NGOs abroad working in a variety of issue areas---human rights, environment, emergency response, and refugee relief.[^not-elite] More importantly, it allows us to directly compare the relative impact of host government restriction of NGOs and organizational financial transparency and accountability on individual donor preferences. The result of this experiment shows that both financial transparency and accountability increase individuals' willingness to donate to NGOs. Conversely, learning about host government criticism of and restrictions on NGOs decreases the likelihood of donation by itself; however, organizational features such as financial transparency and accountability can protect against this dampening effect.

@@

Our findings make two main contributions. First, existing research on privately-provided foreign aid primarily looks at organizations like foundations and large donor NGOs, and not private individuals [@ButheMajordeMelloeSouza:2012; @Koch:2009; @DesaiKharas:2018], and studies that do examine individual giving behavior typically only examine wealthy individuals [@USTrust2014]. Our research expands work on private international philanthropy by exploring how non-elite individuals decide to donate to NGOs focused on international affairs. Second, research shows that NGO leaders often have difficulty in attracting new donors and also find it challenging to motivate current donors [@GaudreauCao:2015]. Examining organizational-level factors that make the average donor increase their charitable giving, especially when facing restricted civic space abroad, can help NGOs better frame their funding appeals. Moreover, our results provide guidance to organizations facing increased antagonism by governments abroad and point to possible methods of adapting to worsening or shrinking civic space.

@@

But as INGOs became increasingly powerful actors in global policy circles, they faced a number of criticisms. Many organizations struggle with the conflicting imperatives of being large organizations that can operate efficiently at scale, while maintaining the grassroots contacts and consensus building that contributed to their early successes [@MitchellSchmitzBruno-vanVijfeijken:2020; @Jalali:2008; @Bano:2012; @Rahman:2006]. In response, many nonprofits sought to establish and enhance their accountability, responsiveness, and legitimacy [@edwardshulme; @GourevitchLake:2012]. These actions were a result of many organizational-level factors impact nonprofit efficacy, including organizational structure, leadership, strategy, transparency and accountability. Organizational and governance structures define how decisions are made and who participates in decision making. While some nonprofits follow more democratic governance models---such as expanded membership to Global South countries, devolving more decision-making powers to the regional or country levels, others have a more centralized structure [@Wong:2012; @MitchellSchmitzBruno-vanVijfeijken:2020, 155]. Leadership can be defined as how leaders "handle the external and internal politics of making their organizations more effective and responsive," and what sort of organizational cultures they create [@MitchellSchmitzBruno-vanVijfeijken:2020, 178]. Finally, strategy is whether the organization has a long-term strategic plan that emphasizes sustainability, innovation, learning and adaptation [@MitchellSchmitzBruno-vanVijfeijken:2020, 100].

However, organizational structure, leadership, and strategy (among other organizational-level factors) are often not visible to the average individual (non-elite) donor, especially in the context of international NGOs. Therefore, in this paper, we focus on the two organizational-level factors that are most visible to the average donor---financial transparency and accountability. Assessing an NGO's deservingness and efficiency demands substantial time and resources, which individual donors typically lack [@CrosonShang:2011; @Tremblay-BoirePrakash:2017]. As such, these donors rely on various easily available cues and heuristics when deciding to support an NGO. News about financial mismanagement and lack of transparency and accountability may be more easily available to these donors. Many nonprofit scandals involve some form of financial wrongdoing [@PrakashGugerty:2010b]. Notable examples include the Oil-for-Food program, initiated by the United Nations in 1996 to allow Iraq to sell enough oil to pay for food and other necessities for its population, which has been suffering from sanctions imposed on Iraq since the First Gulf War. However, in 2004, the widespread financial fraud by the program, limited distribution of food by nonprofits and lack of accountability was widely publicized. Numerous other such cases---covered in @GibelmanGelman:2001 and @GibelmanGelman:2004---show how financial mismanagement, embezzlement, fraud, and lack of accountability can decrease public trust in and donations towards these nonprofits. Globally, around 5% of revenue is lost annually due to fraud [@ACFE:2021]. Misappropriation of assets and embezzlement schemes are among the most common forms of financial fraud [@LamotheTer-MkrtchyanRuddle:2023]. To save time and present easily accessible information, watch dogs such as Charity Navigator and Better Business Bureau Wise Giving Alliance often rate charities on financial and accountability metrics and present this information in a quick, user-friendly format.

@@

We first examine the impact of two organizational practices---financial transparency and accountability---on individual donor preferences. Financial transparency, or the "degree of completeness of information provided by [organizations] to the [public] concerning [their] activities," [@VaccaroMadsen:2006, p. 146] is one tangible heuristic that donors turn to when making decisions. Transparency can involve proactive sharing of information with stakeholders, participation in data transparency platforms, avoiding financial misconduct, and exhibiting trustworthiness. NGOs can engage in financial transparency in a variety of ways such as distribution of audited financial statements and voluntarily sharing information with third party intermediaries. Research shows that transparency is associated with greater donations, lower debt, and better governance [@HarrisNeely:2021; @HarrisPetrovitsYetman:2023; @SaxtonKuoHo:2012]. It can also enhance donor trust and confidence in organizations [@Vaccaro:2006; @VaccaroMadsen:2006].

Previous research has found that media exposés about charity mismanagement can generate negative reputational spillovers for the charitable sector [@GibelmanGelman:2004]. Diversions, or the unauthorized use of an organization's assets, including embezzlement and theft can result in a decrease in donations. This effect becomes even stronger with media coverage of diversion [@HarrisPetrovitsYetman:2023]. Organizational financial transparency can reduce actual or perceived information asymmetries between donors and charities, thus potentially increasing an individual's likelihood of donating to an INGO.

Accountability is another factor that can impact donor preferences. Accountability is the willingness of the organization to explain its action to shareholders [@CharityNavigator:2020] and build and manage relationships with stakeholders or client populations and donor groups. It also encompasses how well an organization's mission is aligned with its resources [@Sloan:2009]. As NGOs gain more prominence, stakeholders, or "any group or individual who can affect or is affected by the achievement of an organization's objectives," [@MitchellAgleWood:1997, 854] want to be assured that their donations or support or support of an organization will make a real difference. More effective accountability can encourage NGOs to become more closely aligned with community practices, provide greater assurances to stakeholders and help expand their support [@Murtaza:2012]. Problems with accountability usually occur when organizations ignore their stakeholders' preferences and use judgements or actions which may not align with one or more of their constituent groups [@McCambridge:2019]. The absence of accountability is associated with decreased public trust, worse reputation, and lower perceived quality [@Becker:2018].

@@

While NGOs have control over organizational factors such as transparency and accountability and can work to improve these characteristics and public perception of them, NGOs may have little to no control over structural factors that can also influence philanthropy. These structural factors include political contexts in NGO host countries, especially a restrictive environment for civil society organizations. A global cascade of anti-NGO laws in recent decades has created barriers to entry, funding, and advocacy for civic organizations [@GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020; @Chaudhry:2022; @Heiss:2017; @DupuyRonPrakash:2016]. The use of such anti-NGO laws, also referred to as administrative crackdown [@Chaudhry:2022], is different from regulations that simply set standards for appropriate organizational behavior and set penalties for violations [@North:1990].[^violent-crackdown] However, the anti-NGO laws discussed in this paper, such as Russian Foreign Agent law, are not grounded in international legal principles, and do not give nonprofits the freedom of association or the freedom to freely access human and material resources [@Voule:2023; @Robinson:2024]. As a result, NGOs must spend more time, effort, and resources to ensure their survival---at the expense of pursuing their missions.

@@

However, previous literature has established that individual private donors do not make the same considerations as official donors when deciding to donate [@DesaiKharas:2018; @ChaudhryHeiss:2021]. Private foreign aid does not have to deal with the same strategic and political considerations as official aid and therefore, may be in a better position to respond to recipients' needs on the grounds [@EasterlyWilliamson:2011]. When a potential recipient NGO faces legal crackdown in its host country, private individual donors can respond by (1) increasing their donations as a sign of support and solidarity, (2) decreasing their donations, punishing the NGO for doing something to run afoul of its host government, or (3) not considering the host country legal environment and making no change in their donation behavior. @ChaudhryHeiss:2021 find conflicting donor responses to foreign restrictions---while some prospective donors punish nonprofits for doing something "wrong" to incur host government restrictions, most tended to increase their support, stating that restricted NGOs are "doing good work in countries where it is tough for groups like them to operate and they need all the help they can get" [@ChaudhryHeiss:2021, 496]. @ChaudhryDotsonHeiss:2021 explore how different combinations of donor characteristics determine which response individuals are likely to take when NGOs face crackdown. They find that donors with longer experiences with the nonprofit sector and high levels of social trust---i.e. those who frequently volunteer, regularly donate to charity, and trust political institutions---are more likely to maintain their support for international NGOs that face criticism or crackdown abroad. Thus, anti-NGO laws may act as a heuristic to individuals donors that NGOs undertake crucial work abroad, which is why governments perceive them as threatening and seek to restrict their work.

This is especially true for nonprofits that rely more heavily on private funds, rather than government funds. Previous research finds that, "individual donors thus seem to be more willing to support besieged human rights organizations when they are unencumbered by government funds." [@ChaudhryHeiss:2021, 496]. Similar trends can be seen even after nonprofits face restrictive political and civic environments in other specific issue areas. For instance, in 2022, after the Supreme Court overturned Roe v. Wade, private philanthropy towards abortion funds skyrocketed [@PBSNews:2022]---a significant development as abortion funds rely largely on individual donors [@Cohen:2022]. Therefore, we expect that anti-NGO restrictions imposed by host-country governments are likely to boost individual donors' inclination to donate to affected organizations.

@@

Donors show increased willingness to donate to NGOs facing government restrictions/criticism while being financial transparent and accountable. While concerns about NGO mismanagement and lack of financial transparency may signal concerns about an NGO's operations and its ability to abide by the host country's regulatory environment, the presence of such transparency and accountability may instead convince donors that government targeting of NGOs may be ill-intentioned rather than simply a manifestation of financial oversight. Donors may view anti-NGO laws and inhospitable civic environments as a sign of governments looking to restrict groups that seek to keep governments accountable, rather than merely auditing and punishing groups that violate routine regulations.

@@

A conjoint research design also allows us to calculate treatment effects even if respondents do not see every combination of organization attributes, and correspondingly provides a substantial increase in statistical power. In standard factorial experimental designs, participants would ordinarily need to be shown all possible combinations of experimental attributes. In our experiment, we presented respondents with six possible treatments with randomized attributes for each treatment, yielding 576 possible unique combinations of features. Even with respondents answering twelve iterations of the experiment, not every combination was seen. However, as long as all the possible organizational attributes are well randomized and there are no systematic biases toward specific options (i.e. more respondents select the first option because it is the first) or toward earlier iterations of attribute choices (i.e. respondents are more careful and attentive for the first hypothetical organization than the last), we can pool all observations together for specific attributes of interest while marginalizing across all other attributes [@KertzerRenshonYarhi-Milo:2021]. This allows us to (1) estimate the effect of each experimental treatment even if some unique combinations of attributes were unseen, and (2) use a much smaller sample size than would be required in a more traditional factorial design.

@@

After presenting respondents with a set of organizations, we asked them which of the three they would be willing to donate to, along with an option for no selection (see @tbl-example for an example). We then repeated the process eleven more times for a total of twelve randomized iterations of hypothetical combinations of attributes, resulting in 12,192 completed experimental tasks (12 iterations × 1,016 respondents). In order to estimate individual-specific differences using a multilevel model, each respondent had to complete multiple experimental tasks. Having the choice of hypothetical organization indicate donation intention and asking respondents to iteratively make this choice across twelve experimental tasks are both typical of conjoint analysis.

@@

We analyze the results using a multilevel Bayesian multinomial model.[^stan] Our experimental data has a natural hierarchical structure, with 3 questions nested inside 12 separate experimental tasks, nested inside each of the 1,016 respondents, which lends itself to multilevel modeling. Relying on standard OLS regression can discard valuable information about (1) the relationships and covariance between the different combinations of feature levels offered to respondents, and (2) individual-specific differences in how respondents react to different feature levels [@JensenMarbleScheve:2021]. Since it was impossible for every respondent to see every possible all 12,000 experimental tasks, multilevel modeling allows us to pool together information about respondents with similar characteristics facing similar sets of choices. Moreover, using random respondent effects provides natural regularization and shrinkage for our estimates---experimental tasks that happened to appear more often due to chance will be accounted for and their frequency will not bias the overall causal effect. We define our model and priors in @eq-model.

>>>>

---
title: "Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy"
short-title: "Why Donors Donate"
date: "2025-01-22"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
  orcid: 0000-0003-3506-7590
- name: Marc Dotson
  affiliation: Utah State University
  url: https://occasionaldivergences.com/
  email: marc.dotson@usu.edu
  orcid: 0000-0001-7846-3660
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Governments across the world have increasingly used laws to restrict the work of nonprofits, which has led to a reduction in public or official foreign aid directed towards these groups. Many international nonprofits, in response, have turned to individual donors to offset the loss of traditional funding. What are individual donors' preferences regarding donating to legally besieged nonprofits abroad? We conducted a conjoint experiment on a nationally representative sample of likely donors in the US and found that learning about host government criticism and legal restrictions on nonprofits decreases individuals' preferences to donate to them. However, organizational features such as financial transparency and accountability can protect against this dampening effect. Our results have important implications both for understanding private international philanthropy and how nonprofits can better frame their fundraising appeals at a time when they are facing restrictive civic spaces and hostile governments abroad.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/andrewheiss/silent-skywalk>
keywords:
  - philanthropy
  - conjoint experiments
  - donor heuristics
  - repression
  - NGOs
  - civil society
  - nonprofits
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/silent-skywalk>"
bibliography: references.bib
csl: csl/apa.csl
link-citations: true
word-count: "9,300"
---

@@

Non-governmental organizations (NGOs) are a core component of global governance and service delivery. Over the last two decades, however, NGOs face restrictive civic spaces abroad [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020]. Repressive governments have increasingly used legal measures to create barriers to funding and advocacy for a variety of NGOs [@DupuyRonPrakash:2016; @Chaudhry:2022; @Heiss:2017]. For instance, in 2012 Russia passed its infamous "Foreign Agent" law (Federal Law No. 121-F2), which forced organizations receiving foreign funding and engaging in political activities to register as foreign agents, which forced hundreds of organizations to close down [@Kirova:2024]. These developments are not limited to authoritarian regimes. Hybrid and democratic states have also restricted NGOs' operations within their territories. Between 2014 and 2019, the Indian government banned 14,500 NGOs from receiving money from abroad, including prominent international nonprofits such as the Ford Foundation and Greenpeace [@PressTrustofIndia:2019]. Policymakers have referred to this phenomenon as the "closing of civic space" or "shrinking civic space" [@CarothersBrechenmacher:2014; @Rutzen:2015b]. Data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in more than 60% of countries---117 in total [@CIVICUSMonitor]. This phenomenon of closing civic space has a large deterrent effect on both advocacy *and* service-delivery NGOs [@SpringmanMaleskyRight:2022].

@@

The amount of private foreign aid is not insignificant---from 2000--2010, private giving by US entities towards international causes doubled in real terms [@DesaiKharas:2018, 505]. In 2018, this amounted to \$22.88 billion in individual donations earmarked specifically for international affairs INGOs [@GivingUSA:2019]. This turn to private donors opens INGOs up to new dynamics in seeking out funds. Individual donor behavior is unlike official donor behavior [@DesaiKharas:2018]. They have different preferences---while official donors channel funds to countries with better institutional quality or to reward governmental performance [@Bermeo:2011; @DietrichWright:2015], individual donors are more inclined to donate to NGOs in countries experiencing a humanitarian crisis [@DesaiKharas:2018]. As such, individual donors may have separate metrics of effectiveness or impact that can differ from the benchmarks that large donor agencies rely on, and INGOs may need to adapt their fundraising strategies to a different set of potential donors.

While INGOs have control over organization-level factors that determine their effectiveness, the political environment in the host countries INGOs work in---including the restrictions on civil society that cause many organizations to seek out private funding in the first place---is a structural factor that INGOs can rarely influence on their own. How do individual donors in the US feel about donating to legally besieged NGOs abroad? How do organizational elements of private donor decision-making interact with broader structural limits on INGO fundraising? While there are many organization-level factors that can determine an NGO's impact, including organizational structure, financial transparency, accountability, leadership, and strategy [@MitchellSchmitzBruno-vanVijfeijken:2020], in this paper, we focus on the two factors that are most visible to donors outside of the organizations and on which information is available publicly---financial transparency and accountability.

@@

We fielded a conjoint experiment to a nationally representative sample of 1,016 adults in the US. This experiment allows us to assess individual-level giving preferences among likely donors in the US towards NGOs abroad working in a variety of issue areas---human rights, environment, emergency response, and refugee relief.[^not-elite] More importantly, it allows us to directly compare the relative impact of host government restriction of NGOs and organizational financial transparency and accountability on individual donor preferences. The result of this experiment shows that both financial transparency and accountability increase individuals' willingness to donate to NGOs. Conversely, learning about host government criticism of and restrictions on NGOs decreases the likelihood of donation by itself; however, organizational features such as financial transparency and accountability can protect against this dampening effect.

@@

Our findings make two main contributions. First, existing research on privately-provided foreign aid primarily looks at organizations like foundations and large donor NGOs, and not private individuals [@ButheMajordeMelloeSouza:2012; @Koch:2009; @DesaiKharas:2018], and studies that do examine individual giving behavior typically only examine wealthy individuals [@USTrust2014]. Our research expands work on private international philanthropy by exploring how non-elite individuals decide to donate to NGOs focused on international affairs. Second, research shows that NGO leaders often have difficulty in attracting new donors and also find it challenging to motivate current donors [@GaudreauCao:2015]. Examining organization-level factors that make the average donor increase their charitable giving, especially when facing restricted civic space abroad, can help NGOs better frame their funding appeals. Moreover, our results provide guidance to organizations facing increased antagonism by governments abroad and point to possible methods of adapting to worsening or shrinking civic space.

@@

But as INGOs became increasingly powerful actors in global policy circles, they faced a number of criticisms. Many organizations struggle with the conflicting imperatives of being large organizations that can operate efficiently at scale, while maintaining the grassroots contacts and consensus building that contributed to their early successes [@MitchellSchmitzBruno-vanVijfeijken:2020; @Jalali:2008; @Bano:2012; @Rahman:2006]. As a result, many nonprofits sought to establish and enhance their accountability, responsiveness, and legitimacy [@edwardshulme; @GourevitchLake:2012]. These actions were a result of many organization-level factors that can potentially impact nonprofit efficacy, including organizational and governance structure, leadership, strategy, transparency, and accountability. Organizational and governance structures define how decisions are made and who participates in decision making. While some nonprofits follow more democratic governance models---such as expanded membership to Global South countries and devolving more decision-making powers to the regional or country levels, others have a more centralized structure [@Wong:2012; @MitchellSchmitzBruno-vanVijfeijken:2020, 155]. Leadership can be defined as how leaders "handle the external and internal politics of making their organizations more effective and responsive," and what sort of organizational cultures they create [@MitchellSchmitzBruno-vanVijfeijken:2020, 178]. Finally, strategy involves whether the organization has a long-term strategic plan that emphasizes sustainability, innovation, learning, and adaptation [@MitchellSchmitzBruno-vanVijfeijken:2020, 100].

However, organizational structure, leadership, and strategy (among other organization-level factors) are often not visible to the average individual (non-elite) donor, especially in the context of international NGOs. Therefore, in this paper, we focus on the two organization-level factors that are most visible to the average donor---financial transparency and accountability. Assessing an NGO's deservingness and efficiency demands substantial time and resources, which individual donors typically lack [@CrosonShang:2011; @Tremblay-BoirePrakash:2017]. As such, these donors rely on various easily available cues and heuristics when deciding to support an NGO. Compared to other organization-level factors, news about financial mismanagement and lack of transparency and accountability may be more easily available. Many nonprofit scandals involve some form of financial wrongdoing [@PrakashGugerty:2010b]. Notable examples include the Oil-for-Food program, initiated by the United Nations in 1996 to allow Iraq to sell enough oil to pay for food and other necessities for its population, which had been suffering from sanctions imposed on Iraq since the First Gulf War. However, in 2004, the program's widespread financial fraud, limited distribution of food by nonprofits, and lack of accountability was widely publicized. Numerous other such cases---covered in @GibelmanGelman:2001 and @GibelmanGelman:2004---show how financial mismanagement, embezzlement, fraud, and lack of accountability can decrease public trust in and donations to these nonprofits. Globally, around 5% of revenue is lost annually due to fraud [@ACFE:2021]. Misappropriation of assets and embezzlement schemes are among the most common forms of financial fraud [@LamotheTer-MkrtchyanRuddle:2023]. To save time and present easily accessible information, watch dogs such as Charity Navigator and Better Business Bureau Wise Giving Alliance often rate charities on financial and accountability metrics and present this information in a quick, user-friendly format.

@@

We first examine the impact of two organizational practices---financial transparency and accountability---on individual donor preferences. Financial transparency, or the "degree of completeness of information provided by [organizations] to the [public] concerning [their] activities," [@VaccaroMadsen:2006, p. 146] is one tangible heuristic that donors turn to when making decisions. Transparency can involve proactive sharing of information with stakeholders, participation in data transparency platforms, avoiding financial misconduct, and exhibiting trustworthiness. NGOs can engage in financial transparency in a variety of ways, such as distribution of audited financial statements and voluntarily sharing information with third party intermediaries. Research shows that transparency is associated with greater donations, lower debt, and better governance [@HarrisNeely:2021; @HarrisPetrovitsYetman:2023; @SaxtonKuoHo:2012]. It can also enhance donor trust and confidence in organizations [@Vaccaro:2006; @VaccaroMadsen:2006].

Previous research has found that media exposés about charity mismanagement can generate negative reputational spillovers for the charitable sector [@GibelmanGelman:2004]. Diversions---or the unauthorized use of an organization's assets, including embezzlement and theft---can result in a decrease in donations. This effect becomes even stronger with media coverage of diversion [@HarrisPetrovitsYetman:2023]. Organizational financial transparency can reduce actual or perceived information asymmetries between donors and charities, thus potentially increasing an individual's likelihood of donating to an INGO.

Accountability is another factor that can impact donor preferences. Accountability is the willingness of the organization to explain its action to shareholders [@CharityNavigator:2020] and build and manage relationships with stakeholders or client populations and donor groups. It also encompasses how well an organization's mission is aligned with its resources [@Sloan:2009]. As NGOs gain more prominence, stakeholders, or "any group or individual who can affect or is affected by the achievement of an organization's objectives," [@MitchellAgleWood:1997, 854] want to be assured that their donations or support or support of an organization will make a real difference. More effective accountability can encourage NGOs to become more closely aligned with community practices, provide greater assurances to stakeholders, and help expand their support [@Murtaza:2012]. Problems with accountability usually occur when organizations ignore their stakeholders' preferences and use judgements or actions which may not align with one or more of their constituent groups [@McCambridge:2019]. The absence of accountability is associated with decreased public trust, deteriorated reputation, and lower perceived quality [@Becker:2018].

@@

While NGOs have control over organizational factors such as transparency and accountability and can work to improve these characteristics and public perception of them, NGOs may have little to no control over structural factors that can also influence philanthropy. These structural factors include political contexts in NGO host countries, especially a restrictive environment for civil society organizations. A global cascade of anti-NGO laws in recent decades has created barriers to entry, funding, and advocacy for civic organizations [@GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020; @Chaudhry:2022; @Heiss:2017; @DupuyRonPrakash:2016]. The use of such anti-NGO laws, also referred to as administrative crackdown [@Chaudhry:2022], is different from regulations that simply set standards for appropriate organizational behavior and set penalties for violations [@North:1990].[^violent-crackdown] However, the anti-NGO laws discussed in this paper, such as the Russian Foreign Agent law, are not grounded in international legal principles and do not give nonprofits the freedom of association or access to resources [@Voule:2023; @Robinson:2024]. As a result, NGOs must spend more time, effort, and resources to ensure their survival---at the expense of pursuing their missions.

@@

However, previous literature has established that individual private donors do not make the same considerations as official donors when deciding to donate [@DesaiKharas:2018; @ChaudhryHeiss:2021]. Private foreign aid does not have to deal with the same strategic and political considerations as official aid and therefore, may be in a better position to respond to recipients' needs on the ground [@EasterlyWilliamson:2011]. When a potential recipient NGO faces legal crackdown in its host country, private individual donors can respond by (1) increasing their donations as a sign of support and solidarity, (2) decreasing their donations, punishing the NGO for perhaps doing something to run afoul of its host government, or (3) not considering the host country legal environment and making no change in their donation behavior. @ChaudhryHeiss:2021 find conflicting donor responses to foreign restrictions---while some prospective donors punish nonprofits for doing something "wrong" to incur host government restrictions, most tended to increase their support, stating that restricted NGOs are "doing good work in countries where it is tough for groups like them to operate and they need all the help they can get" [@ChaudhryHeiss:2021, 496]. @ChaudhryDotsonHeiss:2021 explore how different combinations of donor characteristics determine which response individuals are likely to take when NGOs face crackdown. They find that donors with longer experiences with the nonprofit sector and high levels of social trust---i.e., those who frequently volunteer, regularly donate to charity, and trust political institutions---are more likely to maintain their support for international NGOs that face criticism or crackdown abroad. Thus, anti-NGO laws may act as a heuristic to individuals donors that NGOs undertake crucial work abroad, which is why governments perceive them as threatening and seek to restrict their work.

This is especially true for nonprofits that rely more heavily on private funds, rather than government funds. Previous research finds that, "individual donors thus seem to be more willing to support besieged human rights organizations when they are unencumbered by government funds" [@ChaudhryHeiss:2021, 496]. Similar trends can be seen even after nonprofits face restrictive political and civic environments in other specific issue areas. For instance, in 2022, after the Supreme Court overturned Roe v. Wade, private philanthropy towards abortion funds skyrocketed [@PBSNews:2022]---a significant development as abortion funds rely largely on individual donors [@Cohen:2022]. Therefore, we expect that anti-NGO restrictions imposed by host-country governments are likely to boost individual donors' inclination to donate to affected organizations.

@@

Donors show increased willingness to donate to NGOs facing government restrictions/criticism while being financially transparent and accountable. While concerns about NGO mismanagement and lack of financial transparency may signal concerns about an NGO's operations and its ability to abide by the host country's regulatory environment, the presence of such transparency and accountability may instead convince donors that government targeting of NGOs may be ill-intentioned rather than simply a manifestation of financial oversight. Donors may view anti-NGO laws and inhospitable civic environments as a sign of governments looking to restrict groups that seek to keep governments accountable, rather than merely auditing and punishing groups that violate routine regulations.

@@

A conjoint research design also allows us to calculate treatment effects even if respondents do not see every combination of organization attributes, and correspondingly provides a substantial increase in statistical power. In standard factorial experimental designs, participants would ordinarily need to be shown all possible combinations of experimental attributes. In our experiment, we presented respondents with six possible treatments with randomized attributes for each treatment, yielding 576 possible unique combinations of features. Even with respondents answering twelve iterations of the experiment, not every combination was seen. However, as long as all the possible organizational attributes are well randomized and there are no systematic biases toward specific options (i.e., more respondents select the first option because it is the first) or toward earlier iterations of attribute choices (i.e., respondents are more careful and attentive for the first hypothetical organization than the last), we can pool all observations together for specific attributes of interest while marginalizing across all other attributes [@KertzerRenshonYarhi-Milo:2021]. This allows us to (1) estimate the effect of each experimental treatment even if some unique combinations of attributes were unseen, and (2) use a much smaller sample size than would be required in a more traditional factorial design.

@@

After presenting respondents with a set of three randomly generated organizations, we asked them which of the three they would be willing to donate to, along with an option for no selection (see @tbl-example for an example).This choice indicated donation intention and reflected the option the respondent preferred most within that set. We then repeated the process eleven more times for a total of twelve randomized iterations of hypothetical combinations of attributes, resulting in 12,192 completed experimental tasks (12 iterations × 1,016 respondents). This iterative process across twelve experimental tasks is typical of conjoint analysis and allows us to estimate individual-level differences using a multilevel model and determine both aggregate respondent preferences and the causal effect of each experimental feature on those preferences.

@@

We analyze the results using a multilevel Bayesian multinomial model.[^stan] Our experimental data has a natural hierarchical structure, with 3 questions nested inside 12 separate experimental tasks, nested inside each of the 1,016 respondents, which lends itself to multilevel modeling. Relying on standard OLS regression can discard valuable information about (1) the relationships and covariance between the different combinations of feature levels offered to respondents, and (2) individual-specific differences in how respondents react to different feature levels [@JensenMarbleScheve:2021]. Since it was impossible for every respondent to see every possible combination of all 12,000 experimental tasks, multilevel modeling allows us to pool together information about respondents with similar characteristics facing similar sets of choices. Moreover, using random respondent effects provides natural regularization and shrinkage for our estimates---experimental tasks that happened to appear more often due to chance will be accounted for and their frequency will not bias the overall causal effect. We define our model and priors in @eq-model.

[^stan]: We use Stan [@Stan-2-32-2 v2.32.2; @cmdstan-2-36 v2.36] through R [@rproject-4-4-2 v4.4.2] and {brms} [@Burkner:2017 v2.22] to estimate the model. We simulate 4 MCMC chains with 5,000 draws in each chain, 1,000 of which are used for warmup, resulting in 16,000 (4,000 × 4) draws per model parameter. We assess convergence with visual inspection, and all chains converge. Complete results from all the models, along with posterior predictive checks, goodness-of-fit measures, and model diagnostics---as well as our code and data---are available `ANONYMIZED_URL`.
